Multiplication, division and factorization\\of p-adic polynomials

### 2015-09-25 polynomials/muldivfact.tex
New paper of p-adic polynomials (empty for now)

\permission{%
Permission to make digital or hard copies of all or part of this work for
personal or classroom use is granted without fee provided that copies
are not made or distributed for profit or commercial advantage and
that copies bear this notice and the full citation on the first page.
Copyrights for components of this work owned by others than the
author(s) must be honored. Abstracting with credit is permitted. To
copy otherwise, or republish, to post on servers or to redistribute to
lists, requires prior specific permission and/or a fee.
Request permissions from Permissions@acm.org.}
\conferenceinfo{ISSAC'15,}{July 6--9, 2015, Bath, United Kingdom.}
\copyrightetc{Copyright \copyright~2015 ACM \the\acmcopyr}
\crdata{978-1-4503-3435-8/15/07\ ...\$15.00\\
DOI: http://dx.doi.org/10.1145/2755996.2756655}

@@

\title{p-Adic Stability In Linear Algebra}

@@

\begin{abstract}
Using the differential precision methods developed previously by the same authors,
we study the $p$-adic stability of standard operations on matrices and vector 
spaces. We demonstrate that lattice-based methods surpass naive methods in many
applications, such as matrix multiplication and sums and intersections of subspaces.
We also analyze determinants, characteristic polynomials and LU factorization using these differential methods.
We supplement our observations with numerical experiments.
\end{abstract}

\category{I.1.2}{Computing Methodologies}{Symbolic and Algebraic Manipulation -- \emph{Algebraic Algorithms}}
\terms{Algorithms, Theory}
\keywords{$p$-adic precision; linear algebra; ultrametric analysis}

@@

For about twenty years, the use of $p$-adic methods in symbolic 
computation has been gaining popularity. Such methods were 
used to compute composed products of polynomials 
\cite{boston-gonzalez-perdry-schost:05a}, 
to produce hyperelliptic curves of genus $2$ with complex multiplication 
\cite{gaudry-houtmann-weng-ritzenthaler-kohel:06a},
to compute isogenies between elliptic curves \cite{lercier-sirvent:08a} 
and to count points on varieties using $p$-adic 
cohomology theories (\emph{cf.} \cite{kedlaya:01a,lauder:04a} and many
followers).
However, a general framework allowing a precise study of $p$-adic 
precision --- the main issue encountered when computing 
with $p$-adic numbers --- was designed only recently in 
\cite{caruso-roe-vaccon:14a}.

In \cite{caruso-roe-vaccon:14a}, we advocate the use of lattices
to track the precision of vectors and points on $p$-adic manifolds.
The main result of \emph{loc.~cit.}, recalled in Proposition \ref{prop:precision},
allows for the propagation of precision using differentials.  While
representing precision by lattices carries a high
cost in space and time requirements, the reduced precision loss
can sometimes overwhelm these costs within a larger algorithm.
In this paper, we apply the ideas of \cite{caruso-roe-vaccon:14a} to
certain linear algebraic tasks.

\smallskip

\noindent
{\bf Main Results.}
We give a number of contexts where lattice-based precision methods
outperform the standard coordinate-wise methods.  The two most striking
examples are an analysis of matrix multiplication and of 
sums and intersections of subspaces.  In Proposition \ref{prop:mulmatrix},
we give a formula for the lattice-precision of the product of two matrices.
In Figure \ref{fig:mulmatrix}, we describe the precision loss in multiplying
$n$ matrices, which appears linear in $n$ when using standard methods
and logarithmic in $n$ when using lattice methods.  We also give an
example in Figure \ref{fig:vectorspace} where lattice methods
actually yield an \emph{increase} in precision as the computation
proceeds.

\begin{comment}
The present paper, which may be considered as a continuation of 
\cite{caruso-roe-vaccon:14a}, aims at using the theory of \emph{loc.~cit.} in order to analyze the $p$-adic stability of many standard 
algorithms for computing with basic algebraic structures: numbers, 
matrices, univariate polynomials and vector spaces. For each considered 
algorithm, we will follow the same protocol: 
first, we will compute the theoretical loss of precision of the 
underlying question using the framework of \cite{caruso-roe-vaccon:14a} 
and then compare it to those observed in available implementations in 
\textsc{pari} \cite{pari}, \textsc{magma} \cite{magma} and/or 
\textsc{sage} \cite{sage}.

The conclusion of our analysis is that generic algorithms which are 
designed to work over an arbitrary field are often quite unstable. Very 
roughly, denoting by $D$ the number of divisions performed by the 
algorithm under analysis and by $q$ the cardinality of the residue 
field, it appears that the numbers of lost digits is about $O(\frac D 
q)$ whereas the theoretical loss of precision is closer than $O(\log_q 
D)$ --- and sometimes even $O(1)$. We would like to underline in 
particular that the instability phenomena are much more apparent when 
the residue field is small. It is the reason why we will always take
$K = \Q_2$ is our examples.
\end{comment}

\smallskip

\noindent
{\bf Organization of the paper.}
Section \ref{ssec:latticeprec} recalls the theory of 
precision developed in \cite{caruso-roe-vaccon:14a} and defines
a notion of diffuse precision for comparing to coordinate-wise methods.
In particular, we recall Proposition \ref{prop:precision}, which
allows the computation of precision using differentials.
Proposition \ref{prop:boundLambdaf2} in Section \ref{ssec:boundLambdaf}
is a technical result that will be used to determine the applicability of
Proposition \ref{prop:precision}.

In Section \ref{ssec:mulmatrix}, we analyze matrix multiplication and report
on experiments that demonstrate the utility of lattice-based precision tracking.
Section \ref{ssec:det} finds the precision of the determinant of a
matrix, and Section \ref{ssec:charpoly} applies the resulting formula to
characteristic polynomials.  We define the precision polygon of a matrix, which
gives a lower bound on the precision of the characteristic polynomial.
This polygon lies above the Hodge polygon of the matrix;
we give statistics on the difference and the amount of diffuse precision.
Finally, we apply Proposition \ref{prop:precision} to LU factorization and
describe experiments with lattice-based precision analysis.
Section \ref{ssec:grassgeo} reviews the geometry of Grassmannians,
which we apply in Section \ref{ssec:grassdiff} to differentiating
the direct image, inverse image, sum and intersection maps
between Grassmannians.  We then report in Section \ref{grassimpl}
on tracking the precision of subspace arithmetic in practice.
In the appendix, we give a proof of Proposition \ref{prop:boundLambdaf}.

The code used to make experiments presented in this paper is 
available at \url{https://github.com/CETHop/padicprec}.

\smallskip

\noindent
{\bf Notation.}
Throughout the paper, $K$ will refer to a complete discrete 
valuation field. Usual examples are finite extensions of $\Qp$ and 
Laurent series fields over a field. We denote by $\val : K \twoheadrightarrow \Z \cup 
\{+\infty\}$ the valuation on $K$, by 
$\OK$ the ring of integers and by $\pi \in K$ an element of valuation 
$1$. We let $\Vert \cdot \Vert$ be the norm associated to $\val$.

\medskip

\section{The theory of p-adic precision}
\label{sec:theory}

The aim of this section is to briefly summarize the content of 
\cite{caruso-roe-vaccon:14a} and fill in certain details.

\subsection{Lattices as precision data}
\label{ssec:latticeprec}

In \cite{caruso-roe-vaccon:14a}, we suggest the 
use of lattices to represent the precision of elements in 
$K$-vector spaces.  We shall contrast this approach with the
\emph{coordinate-wise method} used in Sage, % and Magma,
where the precision of an element is specified by giving the precision
of each coordinate separately and is updated after each basic
operation.

Consider a finite 
dimensional\footnote{The framework of \cite{caruso-roe-vaccon:14a} is 
actually those of Banach spaces. However, we will not need infinite 
dimensional spaces here.} normed vector space $E$ 
defined over $K$. We use the notation $\Vert \cdot \Vert_E$ for the norm 
on $E$ and $B^-_E(r)$ (resp. $B^{\phantom -}_E(r)$) for the open (resp. 
closed) ball of radius $r$ centered at the origin. A \emph{lattice} $L \subset 
E$ is a sub-$\O_K$-module which generates $E$ over $K$. 
Since we are working in a ultrametric world, the balls $B^{\phantom 
-}_E(r)$ and $B^-_E(r)$ are examples of lattices. Actually, lattices 
should be thought of as special neighborhoods of $0$ and therefore are good 
candidates to model precision data. Moreover, as revealed in 
\cite{caruso-roe-vaccon:14a}, they behave quite well under (strictly) 
differentiable maps:

\begin{prop}
\label{prop:precision}
Let $E$ and $F$ be two finite dimensional normed vector spaces over $K$ 
and $f : U \rightarrow F$ be a function defined on an open subset $U$ of 
$E$. We assume that $f$ is differentiable at some point $v_0 \in U$ and 
that the differential $f'(v_0)$ is surjective.
Then, for all $\rho \in (0, 1]$, there exists a positive real
number $\delta$ such that, for all $r \in (0, \delta)$, any lattice
$H$ such that $B^-_E(\rho r) \subset H \subset B^{\phantom -}_E(r)$ 
satisfies:
\begin{equation}
\label{eq:firstorder}
f(v_0 + H) = f(v_0) + f'(v_0) (H).
\end{equation}
\end{prop}

This proposition enables the \emph{lattice method} of tracking precision,
where the precision of the input is specified as a lattice $H$ and precision
is tracked via differentials of the steps within a given algorithm.
The equality sign in Eq.~\eqref{eq:firstorder} shows that this method
yields the optimum possible precision. 
We refer to \cite[\S 4.1]{caruso-roe-vaccon:14a} for a more complete 
exposition.

In \cite{caruso-roe-vaccon:14a}, we also explained that if 
$f$ is locally analytic, then the constant $\delta$ appearing in
Proposition \ref{prop:precision} can be expressed in terms of the
growing function $\Lambda(f)$ defined by
\[
\textstyle \Lambda(f)(v) = 
\log \big( \sup_{h \in B^-_E(e^v)} \Vert 
f(h) \Vert \big)
\]
with the convention that $\Lambda(f)(v) = +\infty$ if $f$ does not
converge on $B^-_E(e^v)$.
We refer to \cite[Proposition 3.12]{caruso-roe-vaccon:14a} for the
precise statement.  We state here the case of
integral polynomial functions. A function $f : E \to F$ is said to be
\emph{integral polynomial} if it is given by multivariate polynomial 
functions with coefficients in $\O_K$ in any (equivalently all) system 
of coordinates associated to a $\OK$-basis of $B_E(1)$.

\begin{prop}
\label{prop:precision2}
We keep the notation of Proposition \ref{prop:precision} and assume 
in addition that $f$ is integral polynomial. Let $C$ be a positive real
number such that $B_F(1) \subset f'(v_0)(B_E(C))$. 
Then Proposition \ref{prop:precision} holds with $\delta = C \cdot
\rho^{-1}$.
\end{prop}

In \cite[Appendix A]{caruso-roe-vaccon:14a}, the theory is extended 
to manifolds over $K$, where the precision 
datum at some point $x$ is a lattice in the tangent space at $x$. 
Propositions \ref{prop:precision} and \ref{prop:precision2} have
analogues obtained by working in charts.  We use this extension
to compute with vector spaces in \S \ref{sec:vectorspaces}.

We will use the following definition in contrasting lattice and coordinate-wise methods. 
Suppose $E$ is equipped with a basis $(e_1, \ldots, e_n)$ and write $\pi_i : E \to Ke_i$ for the projections.

\begin{deftn} \label{def:diffused}
Let $H \subset E$ be a lattice. The number of 
\emph{diffused digits of precision} of $H$ is the length of
$H_0/H$ where $H_0 = \pi_1(H) \oplus \cdots \oplus \pi_n(H)$.
\end{deftn}
If $H$ represents the actual precision of some object, then
$H_0$ is the smallest diagonal lattice containing $H$.  Since
coordinate-wise methods cannot yield a precision better than $H_0$,
$k$ provides a lower bound on the number of $p$-adic digits
gained by lattice methods over standard methods.

\subsection{A bound on a growing function}
\label{ssec:boundLambdaf}

In the next sections, we will compute the derivative of several standard 
operations and sometimes give a simple expression in term 
of the input and the output. In other words, the function $f$ modeling
such an operation satisfies a differential equation of the form
$f' = g \circ (f, \id)$  %%%%%%%% HERE
where $g$ is a given --- and hopefully rather simple --- function. The 
aim of this subsection is to study this differential equation and to 
derive from it certain bounds on the growing function $\Lambda(f)$. 
\emph{We will assume that $K$ has 
characteristic $0$.}

Let $E, F$ and $G$ be finite-dimensional normed vector spaces with
$U \subseteq E$ and $V \subseteq F$ and $W \subset G$ open subsets.
Generalizing the setting above, we consider the 
differential equation:
\begin{equation}
\label{eq:diffequah}
f' = g \circ (f, h).
\end{equation}
Here $g : V \times W \to \Hom(E, F)$ and $h : U \to W$ are known
locally analytic functions and $f : U \to V$ is the 
unknown locally analytic function. 
In what follows, we always assume that $V$ and $W$ contain the origin, 
$f(0) = 0$, $h(0) = 0$ and $g(0) \neq 0$. These assumptions are harmless 
for two reasons: first, we can always shift $f$ and $h$ (and $g$ 
accordingly) so that they both vanish at $0$, and second, in order to 
apply Proposition \ref{prop:precision2} the derivative $f'(0)$ 
needs to be surjective and therefore \emph{a fortiori} nonzero.

We assume that we are given in addition two nondecreasing convex 
functions $\Lambda_g$ and $\Lambda_h$ such that $\Lambda(g) \leq 
\Lambda_g$ and $\Lambda(h) \leq \Lambda_h$. We suppose further that 
there exists $\nu$ such that $\Lambda_g$ is constant on the interval 
$({-}\infty, \nu]$\footnote{We note that this assumption is fulfilled if 
we take $\Lambda_g = \Lambda(g)$ because we have assumed that $g(0)$ 
does not vanish.}. We introduce the functions $\tau_\nu$ and $\Lambda_f$ 
defined by:

\vspace{-0.2cm}

\begin{align*}
\tau_\nu(x) &= \smash{\left\{\begin{array}{ll} x & \mbox{ if } x \le \nu, \\
+\infty & \mbox{ otherwise;} \end{array}\right.} \\[8pt]
\mbox{and } \Lambda_f(x) &= \tau_\nu \circ (\id + \Lambda_g \circ \Lambda_h)(x + \alpha),
\end{align*}
%$$\begin{array}{rll}
%&\tau_\nu(x) = x & \text{if } x \leq \nu \\
%& \hphantom{\tau_\nu(x)} = {+} \infty & \text{otherwise} \medskip \\
%\text{and} &
%\multicolumn{2}{l}{\Lambda_f(x) = 
%  \tau_\nu \circ (\id + \Lambda_g \circ \Lambda_h)(x + \alpha)}
%\end{array}$$
where $\alpha$ is a real number satisfying $\Vert n! \Vert \geq 
e^{-\alpha n}$ for all $n$. If $p$ is the characteristic of the residue field,
a suitable value for $\alpha$ is $\alpha = - 
\frac p {p-1} \cdot \log \Vert p \Vert$ if $p > 0$ and $\alpha = 0$ if $p = 0$.
The next Proposition is proved in Appendix \ref{app:proof}.

\begin{prop}
\label{prop:boundLambdaf} \label{PROP:BOUNDLAMBDAF}
We have $\Lambda(f) \leq \Lambda_f$.
\end{prop}

\begin{figure}
\null \hfill
\begin{tikzpicture}[scale=0.8]
\draw[thick,green!30,fill=green!20] 
   (-2,-1)--(-1,2)--(-1,3)--(3.5,3)
 --(2.5,2)--(-3.5,-4)--cycle;
\draw[->] (-5,0)--(3.5,0);
\draw[->] (0,-5)--(0,3);
\node[below, scale=0.75] at (3.5,0) { $x$ };
\node[right, scale=0.75] at (0,3) { $y$ };
\draw[black!80,dotted] (-1,2)--(0,2);
\node[right,scale=0.75] at (0,2) { $\nu$ };
\draw[black!80,dotted] (-2,0)--(-2,-1)--(0,-1);
\node[above,scale=0.75] at (-2.4,0) { $\Lambda_h^{-1}(\nu){-}\alpha$ };
\node[right,scale=0.75] at (0,-1) { $\Lambda_h^{-1}(\nu){+}\mu$ };
\draw[Blue,very thick] (-5,-4)--(-2,-1)--(-1,2);
\fill[Blue] (-1,2) circle (0.06);
\draw[Blue] (-1,2)--(-1,3);
\draw[Sepia, very thick] (-4.5,-5)--(3.5,3);
\draw[Blue, very thick,dashed] (-2,-1)--(-3.5,-4);
\node[below right,rotate=45,scale=0.9] at (-4.3,-4.8) 
  { $y = x + \log \Vert g(0) \Vert$ };
\node[above right,rotate=45,scale=0.9] at (-4.8,-3.8) 
  { $y = x + \alpha + \mu$ };
\end{tikzpicture}
\hfill \null

\vspace{-0.1cm}

\caption{Admissible region for the graph of $\Lambda(f)$}
\label{fig:area}

\end{figure}

Figure \ref{fig:area} illustrates Proposition \ref{prop:boundLambdaf}. 
The blue plain line represents the graph of the function $\Lambda_f$. A 
quick computation shows that, on a neighborhood of ${-}\infty$, this 
function is given by $\Lambda_f(x) = x + \alpha + \mu$
where $\mu$ is the value that $\Lambda_g$ takes on the interval 
$({-}\infty, \nu]$. Proposition \ref{prop:boundLambdaf} says that the
graph of $\Lambda(f)$ lies below the plain blue line. Moreover, we remark
that the Taylor expansion of $f(z)$ starts with the term
$g(0) z$. Hence, on a neighborhood of ${-}\infty$, we have 
$\Lambda(f)(x) = x + \log \Vert g(0) \Vert$. Using convexity, we 
get:
$\Lambda(f)(x) \geq x + \log \Vert g(0) \Vert$
%  \quad \forall x \in \R.$$
for all $x \in \R$.  In other words, the graph of $\Lambda(f)$ lies above the brown line.
Furthermore, we know that the slopes of $\Lambda(f)$ are all integral
because $f$ is locally analytic. Hence, $\Lambda(f)$ cannot lie above
the dashed blue line defined as the line of slope $2$ passing through
the first break point of the blue plain line, which has coordinates
$(y_0 - \alpha - \mu, y_0)$ with $y_0 = \min(\Lambda_h^{-1}(\nu) + \mu, 
\nu)$. As a conclusion, we have proved that the graph of $\Lambda(f)$ 
must coincide with the brown line until it meets the dashed blue line 
and then has to stay in the green area.

As a consequence, we derive the following 
proposition, which can be combined with Proposition 3.12 of 
\cite{caruso-roe-vaccon:14a}. Following \cite{caruso-roe-vaccon:14a},
if $\varphi$ is a convex function and $v \in \R$, we define
$$\varphi_{\geq v} : x \mapsto \inf_{y \geq 0} \big( \varphi(x+y)
- vy \big).$$
It is the highest convex function with $\varphi_{\geq v} \leq \varphi$ and 
$\varphi'_{\geq v} \geq v$.

\begin{prop}
\label{prop:boundLambdaf2}
Keeping the above notation, we have:
$$\Lambda(f)_{\geq 2} (x) \leq 2(x + \alpha + \mu) -
\min(\Lambda_h^{-1}(\nu) + \mu, \: \nu)$$
for all $x \leq \min(\Lambda_h^{-1}(\nu) - \alpha, \: \nu - \mu - \alpha)$.
\end{prop}

\begin{proof}
The inequality follows from the fact that $y = 2(x + \alpha + \mu) - y_0$ is the equation of 
the dashed blue line.
\end{proof}

\begin{rem}
In certain situations,
it may happen that the function $f$ is solution of a simpler 
differential equation of the form $f' = g \circ f$. If this holds, 
Proposition \ref{prop:boundLambdaf2} gives the bound $\Lambda(f)_{\geq 
2} (x) \leq 2(x + \alpha + \mu) - \nu$ for $x \leq \nu - \mu - \alpha$.

Beyond this particular case, we recommend choosing the 
function $h$ by endowing $F$ with the second norm $\Vert x \Vert'_F = 
e^\mu \cdot \Vert x \Vert_F$ ($x \in F$) and taking $h : (F, \Vert 
\cdot \Vert_F) \to (F, \Vert \cdot \Vert'_F)$ to be the identity on the
underlying vector spaces. The function $\Lambda(h) : \R \to \R$ then maps $x$ to $x + 
\mu$ and we can choose $\Lambda_h = \Lambda(h)$.
\end{rem}

\section{Matrices}
\label{sec:matrices}

Let $M_{m,n}(K)$ denote the space of $m \times n$ matrices over $K$.  
We will repeatedly use the Smith decomposition for $M \in M_{m,n}(K)$,
which is $M = U_M \cdot \Delta_M \cdot V_M$
with $U_M$ and $V_M$ unimodular and $\Delta_M$ diagonal.  Write $\sigma_i(M)$
for the valuation of the $(i,i)$-th entry of $\Delta_M$, and by convention set
$\sigma_i(M) = +\infty$ if $i > \min(m,n)$.  Order the $\sigma_i(M)$ so that $\sigma_i(M) \le \sigma_{i+1}(M)$.

\subsection{Multiplication}
\label{ssec:mulmatrix}

To begin with, we want to study the behavior of the precision when 
performing a matrix multiplication. Let $r$, $s$ and $t$ be three 
positive integers and assume that we want to multiply a matrix $A \in 
M_{r,s}(K)$ by a matrix $B \in M_{s,t}(K)$. This operation is of course 
modeled by the integral polynomial function:
$$\begin{array}{rcl}
\mathcal P_{r,s,t} : \quad M_{r,s}(K) \times M_{s,t}(K) & \to & 
M_{r,t}(K) \smallskip \\
(A,B) & \mapsto & AB.
\end{array}$$
According to Proposition \ref{prop:precision}, the behavior of the precision when 
computing $AB$ is governed by $\mathcal P'_{r,s,t}(A,B)$, the linear mapping that takes a pair 
$(dA,dB)$ to $A \cdot dB + dA \cdot B$.

To fix ideas, let us assume from now that the entries of $A$ and $B$ all 
lie in $\O_K$ and are known at the same precision $O(\pi^N)$. In order 
to apply Propositions \ref{prop:precision} and \ref{prop:precision2}, we then need to compute the image 
of the standard lattice $\mathcal L_0 = M_{r,s}(\O_K) \times 
M_{s,t}(\O_K)$ under $\mathcal P'_{r,s,t}(A,B)$. It is of course 
contained in $M_{r,t}(\O_K)$; this reflects the obvious fact that each 
entry of the product $AB$ is also known with precision at least $O(\pi^N)$. 
Nonetheless, it may happen that the above inclusion is strict, meaning 
that we are \emph{gaining} precision in those cases.

Set $a_i = \sigma_i(A)$ and $b_i = \sigma_i(B)$, and define $M_{r,t}((a_i),(b_j))$ 
as the sublattice of $M_{r,t}(\O_K)$ consisting of matrices $M = (M_{i,j})$ 
such that $\val(M_{i,j}) \geq \min(a_i,b_j)$ for all $(i,j)$.  
%Write $a^{(m)}$ (resp. $b^{(m)}$)
%for the number of $a_i$'s (resp. $b_i$'s) which are at least $m$.

\begin{prop}
\label{prop:mulmatrix}
With the above notation, we have
\[
\begin{array}{rl}
& \mathcal P'_{r,s,t}(A,B)(\mathcal L_0)
= U_A \cdot M_{r,t}((a_i),(b_j)) \cdot V_B \smallskip \\
\text{and} &
\text{\emph{length}}\Big(\frac{M_{r,t}(\O_K)}{\mathcal P'_{r,s,t}(A,B)(\mathcal L_0) }
\Big) =
\displaystyle 
\sum_{i,j} \min(a_i,b_j)
\end{array}
\]
\end{prop}

\begin{proof}
We write $A \cdot dB + dA \cdot B = U_A \cdot M \cdot V_B$ with
$$M = \Delta_A \cdot V_A \cdot dA \cdot V_B^{-1} 
+ U_A^{-1} \cdot dB \cdot U_B \cdot \Delta_B.$$
When $dA$ varies in $M_{a,b}(\O_K)$ so does $V_A \cdot dA \cdot V_B^{-1}$
and therefore the first summand in $M$ varies in the subspace of 
$M_{r,t}(\O_K)$ consisting of matrices whose entries on the $i$-th
row have valuation at least $a_i$. Arguing similarly for the second
summand, we deduce the first statement of the Proposition. The second
statement is now clear.
\end{proof}

From the perspective of precision, the second statement of Proposition 
\ref{prop:mulmatrix} means that the computation of $AB$ gains 
$\sum_{i,j} \min(a_i,b_j)$ significant digits in absolute 
precision\footnote{We
note that, on the other side, the valuation of the entries of $AB$ may 
increase, meaning that we are also losing some significant digits if we 
are reasoning in relative precision.} as soon as $N > \min(a_r, b_t)$ 
(\emph{cf.} Proposition \ref{prop:precision2}). However, many of these digits are 
diffused in the sense of Definition \ref{def:diffused}.
To change bases in order to make this increased precision visible
with coordinates, write 
$AB = U_A \cdot P \cdot V_B$ with
$P = \Delta_A \cdot V_A \cdot U_B \cdot \Delta_B$.
Tracking precision in the usual way, the $(i,j)$-th entry
of $P$ is known at precision $O(\pi^{N + \min(a_i,b_j)})$.  Multiplication by
$U_A$ and $V_B$ then diffuses the precision across the entries of $AB$.

\smallskip

We now consider the impact of tracking this diffuse precision.
Although the benefit is not substantial for
a single product of matrices, it accumulates as we multiply a large
number of matrices.  We illustrate this phenomenon with the following simple
example.

\noindent\hrulefill

\noindent {\bf Algorithm 1:} {\tt example\_product}

\noindent{\bf Input:} a list $(M_1, \ldots, M_n)$ of square matrices
of size $d$.

\smallskip

\noindent 1.\ Set $P$ to the identity matrix of size $d$

\noindent 2.\ {\bf for} $j=1,\dots,n$ {\bf do} {\bf compute} $P = P M_i$

\noindent 3.\ {\bf return} the top left entry of $P$

\vspace{-1ex}\noindent\hrulefill

\medskip

\noindent
Figure \ref{fig:mulmatrix} compares the number of significant digits in 
\emph{relative} precision we are losing on the output of Algorithm 1 
when we are using, on the one hand, a standard coordinate-wise track of 
precision and, on the other hand, a lattice-based method to handle 
precision.
%
\begin{figure}
\begin{center}
\renewcommand{\arraystretch}{1.2}
\begin{tabular}{|c|c|c|c|}
\hline
\multirow{2}{*}{\hspace{0.2cm}$d$\hspace{0.2cm}} & 
\multirow{2}{*}{\hspace{0.2cm}$n$\hspace{0.2cm}} & 
\multicolumn{2}{|c|}{Average loss of precision} \\
\cline{3-4}
& & Coord-wise method & Lattice method \\
\hline 
$2$ & $10$ & $\hphantom{00}2.8$ & $2.4$ \\
$2$ & $100$ & $\hphantom{0}16.7$ & $5.0$ \\
$2$ & $1000$ & $157.8$ & $7.9$ \\
\hline
$3$ & $10$ & $\hphantom{00}2.2$ & $1.9$ \\
$3$ & $100$ & $\hphantom{0}12.8$ & $4.0$ \\
$3$ & $1000$ & $122.5$ & $7.0$ \\
\hline
\end{tabular}

\smallskip

{\small
Results for a sample of $1000$ random inputs in $M_{d,d}(\Z_2)^n$}
\end{center}
\renewcommand{\arraystretch}{1}

\vspace{-0.3cm}

\caption{Average loss of precision in Algorithm 1}
\label{fig:mulmatrix}
\end{figure}
%
We observe that, in the first case, the number of lost digits seems to 
grow linearly with respect to the number of multiplications we are 
performing (that is $n$) whereas, in the second case, the growth seems 
to be only logarithmic. It would be nice to have a precise formulation 
(and proof) of this heuristic.

Note that multiplication of many random matrices plays a central
role in the theory of random walks on homogeneous spaces \cite{Benoist:2012}.
Better stability in computing such products helps estimate
Lyapunov exponents in that context.

\subsection{Determinant}
\label{ssec:det}

The computation of the differential of $\det : M_{n,n}(K) \to K$
is classical: at a point $M$ it is the linear map
\[
\detp(M) : dM \mapsto \tr(\com(M) \cdot dM),
\]
where $\com(M)$ stands for the comatrix of $M$, which is $\det(M) M^{-1}$ when $M$ is invertible. If $\rank(M) < n - 1$,
then $\detp(M)$ is not surjective and we cannot apply Proposition 
\ref{prop:precision}. Therefore, we suppose that $\rank(M) \ge n-1$ for the rest of this section.

As with matrix multiplication, we first determine the image of
the standard lattice $\mathcal L_0 = M_{n,n}(\O_K)$ under $\detp(M)$.

\begin{prop}
\label{prop:detmatrix}
Setting $v = \sigma_1(M) + \cdots + \sigma_{n-1}(M)$, we have
$\detp(M)(\mathcal L_0) = \pi^v \O_K$.
\end{prop}

\begin{proof}
From the description of $\detp(M)$, we see that it is enough to prove 
that the smallest valuation of an entry of $\com(M)$ is $v$
or, equivalently, that the ideal of $\O_K$ generated by all 
minors of $M$ of size $(n-1)$ is $\pi^v
\O_K$. But this ideal remains unchanged when we multiply $M$ 
on the left or on the right by a unimodular matrix. Thus we may 
assume that $M = \Delta_M$, and the result becomes clear.
\end{proof}

In terms of precision, Proposition \ref{prop:detmatrix} implies that if 
$M$ is given at flat precision $O(\pi^N)$ with $N > v$, then $\det(M)$ 
is known at precision $O(\pi^{N+v})$.  Thus we are gaining $v$  
digits in absolute precision or, equivalently, losing $\sigma_n(M)$ digits
of relative precision.
Furthermore, one may compute $\det(M)$ with this optimal precision by finding an approximate
Smith decomposition with $\Delta_M$ known at precision $O(\pi^N)$ and
multiplying its diagonal entries.

\subsection{Characteristic polynomials} \label{ssec:charpoly}

Write $\charpoly : M_{n,n}(K) \to K[X]$ for the characteristic 
polynomial, and $K_{<n}[X] \subset K[X]$ for the subspace consisting of 
polynomials of degree less than $n$.  Then the differential of 
$\charpoly$ at a point $M$ is
\[
\charp(M) : dM \mapsto \tr(\com(X - M) \cdot dM).
\]
The image is the $K$-span of the entries of $\com(X{-}M)$, which is 
clearly contained within $K_{<n}[X]$.  In fact, the image will equal 
$K_{<n}[X]$ as long as $M$ does not have two Jordan blocks with the same 
generalized eigenvalue. For now on, we make this assumption.

Recall that the \emph{Newton polygon} $\NP(f)$ of a polynomial $f(X) = 
\sum_{i} a_i X^i$ is the lower convex hull of the points $(i, 
\val(a_i))$ and the \emph{Newton polygon} $\NP(M)$ of a matrix $M$ is 
$\NP(\charpoly(M))$.  The \emph{Hodge polygon} $\HP(M)$ of $M$ is the lower 
convex hull of the points $(i, \sum_{j = 1}^{n-i} \sigma_j(M))$. For 
any matrix $M$, the polygon $\NP(M)$ lies above $\HP(M)$ 
\cite{kedlaya:padicDiffEq}*{Thm. 4.3.11}.

Such polygons arise naturally in tracking the precision of polynomials 
\cite{caruso-roe-vaccon:14a}*{\S 4.2}.  Any such polygon $P$ yields a 
lattice $\mathcal{L}_P$ in $K_{<n}[X]$ consisting of polynomials whose 
Newton polygons lie above $P$.  This lattice is generated by monomials 
$a_iX^i$, where $\val(a_i)$ is the ceiling of the height of $P$ at $i$.  
These polygons are used in a coordinate-wise precision tracking for polynomial
arithmetic. We now introduce another polygon, bounded 
between $\NP(M)$ and $\HP(M)$, that will provide an estimate on the precision of $\charpoly(M)$.

\begin{deftn}
The \emph{precision polygon} $\PP(M)$ of $M$ is the lower convex hull of 
the Newton polygon of the entries of $\com(X{-}M)$.
\end{deftn}

It is clear from the definition $\mathcal L_{\PP(M)} \subset 
\charp(M)(\mathcal{L}_0)$ where $\mathcal{L}_0$ is the standard lattice 
$M_{n,n}(\OK)$. More precisely, $\PP(M)$ is the smallest polygon $P$ for 
which the inclusion $\mathcal L_P \subset \charp(M)(\mathcal{L}_0)$ holds. By 
Proposition \ref{prop:precision}, the precision polygon determines the minimal 
precision losses possible when encoding polynomial precision using polygons.

It turns out that the precision polygon is related to the Hodge and Newton 
polygons. If a polygon $P$ has vertices $(x_i, y_i)$, we let $T_n(P)$ be 
the translated polygon with vertices $(x_i - n, y_i)$.

\begin{prop} \label{prop:polygons}
The precision polygon $\PP(M)$ lies between $T_1(\HP(M))$
and $\NP(M)$.

Moreover, $\PP(M)$ and $T_1(\HP(M))$ meet at $0$ and $n{-}1$.
\end{prop}

\begin{proof}
The coefficients of $\charpoly(M)$ can be expressed as traces of 
exterior powers: the coefficient of $X^i$ is $\tr(\Lambda^i(M))$,
which is $\tr(\Lambda^i(U_M) \Lambda^i(\Delta_M) \Lambda^i(V_M))$.  
Computing $\Lambda^i(\Delta_M)$, we get the first statement of
the Proposition. For $i = 1$, we further find that $\PP(M)$ vanishes
at the abscissa $n{-}1$. By definition so does $T_1(\HP(M))$. The fact 
that $\PP(M)$ and $T_1(\HP(M))$ meet at abscissa $0$ follows from
Proposition \ref{prop:detmatrix}.

It remains to prove the comparison with the Newton polygon.
Set $f = \charpoly(M)$, set $m_{i,j}$ as the $(i,j)$-th entry of $M$, 
$f_{i,j}$ as the $(i,j)$-th entry of $\com(X{-}M)$ and $\mu_{i,j} = 
\val(m_{i,j})$. We write $f[k]$ 
for valuation of the coefficient of $X^k$ in $f$, and set $f[-1] = 
+\infty$. The equation $(X{-}M) \cdot \com(X{-}M) = f \cdot I$ yields, 
for all $i$ and $k$,
\begin{align*}
f[k] &\ge \inf(f_{i,i}[k{-}1], \mu_{i,0} + f_{0,i}[k], \ldots, 
\mu_{i,n} + f_{n,i}[k]) \\
&\ge \inf(f_{i,i}[k{-}1], f_{j,i}[k]),
\end{align*}
with the infimum over $j$. 
Taking lower convex hulls and noting that $\PP(M)$ is nonincreasing,
which follows from the comparison with the Hodge polygon, we get the 
result.
\end{proof}

\begin{rem}
Experiments actually support the following stronger result: $\PP(M)$ is 
bounded above by $T_1(\NP(M))$.
\end{rem}

For many matrices, $\PP(M) = T_1(\HP(M))$.  For random matrices over 
$\Z_2$, the $2$-adic precision polygon is equal to the Hodge polygon in 
$99.5\%$ of cases in dimension $4$, down to $99.1\%$ in dimension $8$.  
Over $\Z_3$, the fraction rises to $99.98\%$, with no clear dependence on 
dimension.
Empirically, $\PP(M)$ seems most likely to differ from $T_1(\HP(M))$ at 
$1$, corresponding to the precision of the linear term of the 
characteristic polynomial.

>>>>

\permission{%
}

@@

\title{Multiplication, division and factorization\\of p-adic polynomials}

@@

\begin{abstract}
\end{abstract}

\category{I.1.2}{Computing Methodologies}{Symbolic and Algebraic Manipulation -- \emph{Algebraic Algorithms}}
\terms{Algorithms, Theory}
\keywords{}

@@

\section{Modular multiplication}

\section{Euclidean division}

\section{Slope factorization}
